A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-15 paper.tex

>>>>

\section*{SARS-CoV-2 ARGs: a case study}
% What is SARS-CoV-2 and why is it interesting here? What's the dataset?
% What are its basic evolutionary properties, summarised for a
% population geneticist?
Recombination via template switching is an important
factor in the evolution of
SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}.
With multiple deeply divergent lineages circulating globally,
coinfection (and the potential for recombination)
is relatively likely~\citep{bal2022detection}.
The COVID-19 pandemic led to collection of whole-genome viral data
at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
currently holding over 15 million sequences [FIXME]. Detection
and characterisation of recombination is of profound interest
for [reasons XYZ].

% Summary of sc2ts paper. We inferred some big ole ARGs, with an
% unprecedented degree of detail about recombination (outside, I guess,
% of agri datasets). Summary of what's previously known about sc2 recomb.
\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
on GISAID data, one containing 1.27 million sequences sampled up to
June 30, 2021 (the ``Wide ARG'', and the second
(more sparsely sampled) consisting of 657K sequences sampled up to June 30,
2022 (the ``Long ARG''). Recombination is rare in both ARGs, with $<0.3\%$
of the nodes being inferred to be recombinants (and quite likely
a substantial number of false positives caused by poor data quality).
Nonetheless, this provides thousands of automatically detected
recombination events, the characterisation of which is usually a manual
and labour intensive process. The recombinations events detected
and fully characterised in these large ARGs fit well with previous
which exhaustively investigate specific
recombinants~\citep{jackson2021generation}.
[Not sure what I'm trying to say here;
what does a PopGen person who knows very little about SARS2 need to
know here to given them context?]

% What's the point of this case study? To emphasise that we
% need to systematically represent topological uncertainty in ARGs.
The pandemic scale SARS-CoV-2 dataset is an interesting case study
for a number of reasons.
Firstly, fundamental properties of the data underscore the
need for topological uncertainty through polytomies
and stacked recombination [more concise way of saying this?].
The scale of sequencing effort at the height
of the pandemic means that there were at times tens of thousands
of sequences collected in a single day. This, combined with
the slow rate of mutation, means that we will often have hundreds
of identical sequences collected on the same day.
This makes polytomies an unescapable reality, and a
fully-resolved binary tree is clearly false precision in this case.
At the other extreme, ``unary'' nodes are also a common and
necessary property of such densely sampled data [explain and
give numbers?]
Even with this exquisite level of detail we are
often unable to fully time-resolve recombination events. While
in some cases a recombination event with two parental lineages
can be quite precisely located in space and time (e.g. XA was in Wales
around [when]), in others three or more quite distinct parental
lineages indicate multiple unobserved recombinations.
[JK: I'm just making stuff up here: I'm assuming that there's an
example of a complex X lineages that people haven't figured the
exact history of. Just seeing if this makes a good narrative
before filling in the details.]
While this
may be seen as a limitation of the \texttt{sc2ts} method,
concerted efforts to disentangle the origins of [some example
of a complicated X lineage] have not succeeded in shedding
much extra light, and truthfully representing the uncertainty
around these events seems a more useful approach than
 arbitrary resolution. Encoding the dataset as an eARG,
with a specific event for each recombination breakpoint,
would require such false precision.

An interesting property of the SARS-CoV-2 dataset is that
it highlights the uncertainty in the genomic position of
recombination breakpoints. [Some discussion about why
we can never know where the breakpoint really happened, and
some follow-up to note that this is even more acute in
standard popgen datasets where we're thinking about ancient
recombs].
This is a limitation of the
gARG encoding proposed here, which requires that we specify
exact intervals of inheritance and therefore precise
recombination breakpoints.
An approach in which we can systematically reason about the
uncertainty of inheritance intervals as they propogate
through an ARG
would be an interesting avenue for future research.
